Adversarial Variational Optimization\\ of Non-Differentiable Simulators

### 2018-02-10 colt/avo.tex
colt submission

\documentclass{article}
\usepackage[accepted]{icml2018}
 % If your paper is accepted, change the options for the package
% aistats2018 as follows:
%
%\usepackage[accepted]{aistats2018}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}

\usepackage{bm}
\usepackage{cancel}
\usepackage{bbold}
\usepackage{bm}
\usepackage{slashed}
\usepackage{graphicx}
\usepackage{color}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{tikz}
\usepackage{hyperref}

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

@@

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}

\icmltitlerunning{Adversarial Variational Optimization of Non-Differentiable Simulators}

@@

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

\twocolumn[

\icmltitle{Adversarial Variational Optimization of\\
           Non-Differentiable Simulators}

\begin{icmlauthorlist}
\icmlauthor{Gilles Louppe}{uliege}
\icmlauthor{Kyle Cranmer}{nyu}
\end{icmlauthorlist}

\icmlaffiliation{uliege}{University of Li{\`e}ge, Belgium}
\icmlaffiliation{nyu}{New York University, USA}

\icmlcorrespondingauthor{Gilles Louppe}{g.louppe@uliege.be}
\icmlcorrespondingauthor{Kyle Cranmer}{kyle.cranmer@nyu.edu}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{likelihood-free inference, implicit models, generative adversarial networks}

\vskip 0.3in
]

\printAffiliationsAndNotice{\icmlEqualContribution}

@@

@@

In many fields of science such as particle physics, epidemiology  or
population genetics, computer simulators are used to describe complex data
generation processes. These simulators relate observations $\bfx$ to the
parameters $\bftheta$ of an underlying theory or mechanistic model. In most
cases, these simulators are specified as procedural implementations of forward,
stochastic processes involving latent variables $\bfz$. Rarely do these
simulators admit a tractable density (or likelihood) $p(\bfx | \bftheta)$. The
prevalence and significance of this problem has motivated an active research
effort in so-called \textit{likelihood-free inference} algorithms such as
Approximate Bayesian Computation (ABC) and density estimation-by-comparison
algorithms~\cite{beaumont2002approximate, marjoram2003markov,
sisson2007sequential, sisson2011likelihood, marin2012approximate,
cranmer2015approximating}.

In parallel, with the introduction of variational
auto-encoders~\cite{DBLP:journals/corr/KingmaW13} and generative adversarial
networks~\cite{goodfellow2014generative}, there has been a vibrant research
program around implicit generative models based on neural
networks~\cite{2016arXiv161003483M}.  While some of these models
also do not admit a tractable density, they are all differentiable by construction.
In addition, generative models based on neural networks are highly parametrized and the model
parameters have no obvious interpretation. In contrast, scientific simulators
can be thought of as highly regularized generative models as they typically have
relatively few parameters and they are endowed with some level of
interpretation. In this setting, inference on the model parameters $\bftheta$ is
often of more interest than the latent variables $\bfz$.

In this work, we develop a likelihood-free inference algorithm
for non-differentiable, implicit generative models.
We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\cite{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives.
The objective of the algorithm is to match the marginal distribution of
the synthetic data to the empirical distribution of observations.

@@

Generative adversarial networks (GANs) were first proposed by
\cite{goodfellow2014generative} as a way to build an implicit generative model
capable of producing samples from random noise $\bfz$. More specifically,
a generative model $g(\cdot; \bftheta)$ is pit against an adversarial
classifier $d(\cdot; \bfphi):\mathbb{R}^d \to [0,1]$ with parameters $\bfphi$ and whose antagonistic objective is to recognize real data $\mathbf{x}$
from generated data $\tilde{\mathbf{x}} = g(\bfz; \bftheta)$. Both models $g$ and $d$
are trained simultaneously, in such a way that $g$ learns to fool
its adversary $d$ (which happens when $g$ produces samples comparable to the
observed data), while $d$ continuously adapts to changes in $g$. When $d$ is
trained to optimality before each parameter update of the generator, it can
be shown that the original adversarial learning procedure~\cite{goodfellow2014generative} amounts to minimizing
the Jensen-Shannon divergence $\text{JSD}(p_r(\mathbf{x}) \parallel p(\mathbf{x}|\bftheta))$ between $p_r(\mathbf{x})$ and $p(\mathbf{x}|\bftheta)$.

As explored in \cite{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\cite{2017arXiv170107875A} (WGANs) reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  )$
by replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$, such that
\begin{align}
& W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) \nonumber \\
&= \sup_{\bfphi}  \mathbb{E}_{\tilde{\mathbf{x}}\sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}}; \bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x}; \bfphi)] \nonumber \\
&\triangleq \sup_{\bfphi} \mathcal{L}_W.
\end{align}
Under the WGAN-GP formulation of \cite{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\bfphi$ and $\bftheta$ in order to respectively minimize
\begin{align}
    {\cal L}_d =\,&  \mathcal{L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& - \mathcal{L}_W
\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)$.

@@

Variational optimization~\cite{2012arXiv1212.4507S,staines2013optimization} and evolution strategies~\cite{2011arXiv1106.4487W} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the observation that
\begin{equation}
    \min_{\bftheta} f(\bftheta) \leq \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] = U(\bfpsi),
\end{equation}
where $q(\bftheta|\bfpsi)$ is a proposal distribution with parameters $\bfpsi$ over input values $\bftheta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal distribution is flexible enough, the parameters $\bfpsi$
can be updated to place its mass arbitrarily tight around the optimum $\bftheta^* = \min_{\bftheta \in \Theta} f(\bftheta)$.

Under mild restrictions outlined in  \cite{2012arXiv1212.4507S}, the bound
$U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
trick its gradient can be rewritten as:
\begin{align}\label{eqn:approx-grad}
    \nabla_\bfpsi U(\bfpsi) &= \nabla_\bfpsi \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] \nonumber \\
    %&= \nabla_\bfpsi \int f(\bftheta)  q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
 %   &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
   % &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
    &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\bfpsi \log
q(\bftheta|\bfpsi)$ of the proposal is known and that one can evaluate
$f(\mathbf{\bftheta})$ for any $\bftheta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\bfpsi)$ with stochastic gradient descent (or a variant thereof, robust to noise
and parameter scaling).

@@

Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
variance of the noisy estimator of the gradients may be too large to be useful in many problems.
We use Adam~\cite{2014arXiv1412.6980K}, but note the opportunity to use instead the Natural Evolution Strategy
algorithm~\cite{2011arXiv1106.4487W} or variance reduction techniques as
control variates~\cite{2017arXiv171100123G} or finite differences~\cite{buesingstochastic}.

@@

The variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
the parametrized marginal distribution of the generated data
\begin{equation}\label{eq:marginal_likelihood}
\qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
\end{equation}
We can think of $q(\bfx|\bfpsi)$ as a \textit{variational program} as described
in \cite{2016arXiv161009033R}, though more complicated than a simple
reparametrization of normally distributed noise $\bfz$ through a differentiable
function. In our case, the variational program is a
marginalized, non-differentiable  simulator.  Its density is  intractable;
nevertheless, it can generate samples for $\bfx$ whose expectations are differentiable with
respect to $\bfpsi$.
Operationally, we sample from this marginal model via
\begin{equation}\label{eqn:p_psi}
\mathbf{x} \sim \qxpsi \triangleq \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
\end{equation}

We can view the optimization of $q(\bfx | \bfpsi)$ with respect to $\bfpsi$ through the lens of empirical Bayes, where the data are used to optimize a prior within the family $q(\bftheta|\bfpsi)$. If $\rho$ is the KL distance, $\bfpsi^*$ would correspond to the maximum marginal likelihood estimator advocated by Rubin~\cite{rubin1984}.
When $\rho$ is the Wasserstein distance, $\bfpsi^*$ is referred to as the
minimum Wasserstein estimator (MWE).
When the model is well specified, the MWE
coincides with the true data-generating parameter; however, if the model is
misspecified, the MWE is typically different from the maximum likelihood
estimator (MLE).
Thus, if the simulator $p(\bfx |
\bftheta)$ is misspecified, $q(\bftheta | \bfpsi)$ will attempt to smear it so that the marginal
model $q(\bfx | \bfpsi)$ is closer to $p_r(\bfx)$. However, if the simulator is well specified,
then $q(\bftheta | \bfpsi)$ will concentrate its mass around the true data-generating parameter.

@@

@@

The observed data is sampled from a Poisson with mean $\lambda^* = 7$.
Algorithm~\ref{alg:avo} is run for 500 epochs with mini-batches of size $M=64$
and the following configuration. For the critic $d$, we use a 3-layer MLP with 10
hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
$n_\text{critic}=10$ iterations with a step size $\alpha=0.005$ and $\beta_1=\beta_2=0.05$.
% Its inner first and second moment vectors are reset at
% each outer epoch in order to avoid building momentum in staled directions.
For
estimating $\lambda^*$, we parametrize $\bftheta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
$q(\bftheta|\bfpsi)$ initialized with a mean at $\log(10)$ and unit variance. At
each epoch, parameters $\bfpsi$ are updated by taking one Adam step and the same hyper-parameters as for the critic.
The gradient penalty coefficient is set to
$\lambda=0.1$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=5$.

The top left plot in Figure~\ref{fig:poisson} illustrates the resulting proposal
distributions $q(\bftheta|\bfpsi)$ after AVO.  For
both $\gamma=0$ and $\gamma=5$, the proposal distributions correctly concentrate
their density around the true parameter value $\log(\lambda^*) = 1.94$. Under
the effect of entropic regularization,
the proposal distribution for $\gamma=5$ concentrates its mass tightly,
yielding in this case precise inference.  The top right plot compares the
model distributions to the true distribution.  As theoretically expected from
adversarial training, we see that the resulting distributions align with
the true distribution, with in this case visually slightly better results for the penalized
model.  The bottom plot of Figure~\ref{fig:poisson} shows empirical estimates
of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=5$,
the curves fall towards $0$, which indicates that
$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)}
[d(\tilde{\mathbf{x}};\bfphi)] \approx \mathbb{E}_{\mathbf{x} \sim
p_r(\mathbf{x})} [d(\mathbf{x};\bfphi)]$ and that the critic cannot distinguish
between true and model data. This confirms that adversarial variational optimization
works despite the discreteness of the data and lack of access to the
density $p(\bfx | \bftheta)$ or its gradient.

\begin{figure}[h]
\centering
\includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
\caption{Discrete Poisson model with unknown mean.
 ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after training. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
            the true value $\log(\lambda^*)$. Entropic regularization ($\gamma=5$) results in a tighter density.
 ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parametrizations of the simulator closely reproduce the true distribution.
 ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
 }\label{fig:poisson}
\end{figure}

@@

As a second toy example, we consider a generator producing
5-dimensional continuous data, as originally specified in Section 4.2 of
\cite{cranmer2015approximating}.
% More specifically, we consider the following
% generative process:
% \begin{itemize}
% \item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
% $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
% $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
% $z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
% $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
% $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
% \item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
% semi-positive definite $5 \times 5$ matrix defining a fixed projection
% of $\bfz$ into the observed space.
% \end{itemize}
Again, AVO does not have access to the density or its gradient, only samples from the generative model.
We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored  (mean field) Gaussian
proposal distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component was
initialized with zero mean and unit variance.
Hyper-parameters are set to $M=64$, $n_\text{critic}=10$, $\lambda=0.1$, $\gamma=1$ and
Adam configured with a step size $\alpha=0.005$ and $\beta_1=\beta_2=0.05$.
The architecture of the critic is the same as in the previous example.

@@

This work sits at the intersection of several lines of research related to
likelihood-free inference, approximate Bayesian computation (ABC),
implicit generative models, and variational inference.
Viewed from the literature around implicit generative models based on neural networks,
the proposed method can be considered as a direct adaptation of
generative adversarial networks~\cite{goodfellow2014generative,2017arXiv170104862A} to
non-differentiable simulators using variational optimization~\cite{2012arXiv1212.4507S}.

>>>>

\documentclass[12pt]{colt2018} % Anonymized submission
% \documentclass{colt2017} % Include author names

% The following packages will be automatically loaded:
% amsmath, amssymb, natbib, graphicx, url, algorithm2e

\title[Adversarial Variational Optimization]{Adversarial Variational Optimization\\
                                             of Non-Differentiable Simulators}
\usepackage{times}
 % Use \Name{Author Name} to specify the name.
 % If the surname contains spaces, enclose the surname
 % in braces, e.g. \Name{John {Smith Jones}} similarly
 % if the name has a "von" part, e.g \Name{Jane {de Winter}}.
 % If the first letter in the forenames is a diacritic
 % enclose the diacritic in braces, e.g. \Name{{\'E}louise Smith}

 % Two authors with the same address
  % \coltauthor{\Name{Author Name1} \Email{abc@sample.com}\and
  %  \Name{Author Name2} \Email{xyz@sample.com}\\
  %  \addr Address}

 % Three or more authors with the same address:
 % \coltauthor{\Name{Author Name1} \Email{an1@sample.com}\\
 %  \Name{Author Name2} \Email{an2@sample.com}\\
 %  \Name{Author Name3} \Email{an3@sample.com}\\
 %  \addr Address}

 % Authors with different addresses:
 \coltauthor{\Name{Gilles Louppe} \Email{g.louppe@uliege.be}\\
 \addr University of Li\`{e}ge, Belgium
 \AND
 \Name{Kyle Cranmer} \Email{kyle.cranmer@nyu.edu}\\
 \addr New York University, USA
 }

 % \usepackage{amsfonts}
 % \usepackage{amssymb}
 % \usepackage{amsmath}
 % \usepackage{amsthm}

 \usepackage{bm}
 \usepackage{cancel}
 \usepackage{bbold}
 \usepackage{bm}
 \usepackage{slashed}
 % \usepackage{graphicx}
 \usepackage{color}
 \usepackage{algorithm}
 \usepackage{algpseudocode}
 \usepackage{tikz}
 \usepackage{hyperref}

 % Attempt to make hyperref and algorithmic work together better:
 % \newcommand{\theHalgorithm}{\arabic{algorithm}}

@@

 % \theoremstyle{plain}
 % \newtheorem{theorem}{Theorem}
 % \newtheorem{proposition}[theorem]{Proposition}

@@

\maketitle

@@

\begin{keywords}
List of keywords
\end{keywords}

@@

In many fields of science such as particle physics, epidemiology  or
population genetics, computer simulators are used to describe complex data
generation processes. These simulators relate observations $\bfx$ to the
parameters $\bftheta$ of an underlying theory or mechanistic model. In most
cases, these simulators are specified as procedural implementations of forward,
stochastic processes involving latent variables $\bfz$. Rarely do these
simulators admit a tractable density (or likelihood) $p(\bfx | \bftheta)$. The
prevalence and significance of this problem has motivated an active research
effort in so-called \textit{likelihood-free inference} algorithms such as
Approximate Bayesian Computation (ABC) and density estimation-by-comparison
algorithms~\citep{beaumont2002approximate, marjoram2003markov,
sisson2007sequential, sisson2011likelihood, marin2012approximate,
cranmer2015approximating}.

In parallel, with the introduction of variational
auto-encoders~\citep{DBLP:journals/corr/KingmaW13} and generative adversarial
networks~\citep{goodfellow2014generative}, there has been a vibrant research
program around implicit generative models based on neural
networks~\citep{2016arXiv161003483M}.  While some of these models
also do not admit a tractable density, they are all differentiable by construction.
In addition, generative models based on neural networks are highly parametrized and the model
parameters have no obvious interpretation. In contrast, scientific simulators
can be thought of as highly regularized generative models as they typically have
relatively few parameters and they are endowed with some level of
interpretation. In this setting, inference on the model parameters $\bftheta$ is
often of more interest than the latent variables $\bfz$.

In this work, we develop a likelihood-free inference algorithm
for non-differentiable, implicit generative models.
We adapt the adversarial
training procedure of generative adversarial
networks \citep{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives.
The objective of the algorithm is to match the marginal distribution of
the synthetic data to the empirical distribution of observations.

@@

Generative adversarial networks (GANs) were first proposed by
\citep{goodfellow2014generative} as a way to build an implicit generative model
capable of producing samples from random noise $\bfz$. More specifically,
a generative model $g(\cdot; \bftheta)$ is pit against an adversarial
classifier $d(\cdot; \bfphi):\mathbb{R}^d \to [0,1]$ with parameters $\bfphi$ and whose antagonistic objective is to recognize real data $\mathbf{x}$
from generated data $\tilde{\mathbf{x}} = g(\bfz; \bftheta)$. Both models $g$ and $d$
are trained simultaneously, in such a way that $g$ learns to fool
its adversary $d$ (which happens when $g$ produces samples comparable to the
observed data), while $d$ continuously adapts to changes in $g$. When $d$ is
trained to optimality before each parameter update of the generator, it can
be shown that the original adversarial learning procedure~\citep{goodfellow2014generative} amounts to minimizing
the Jensen-Shannon divergence $\text{JSD}(p_r(\mathbf{x}) \parallel p(\mathbf{x}|\bftheta))$ between $p_r(\mathbf{x})$ and $p(\mathbf{x}|\bftheta)$.

As explored in \citep{2017arXiv170104862A}, GANs remain remarkably
difficult to train because of vanishing gradients as $d$ saturates, or because of
unreliable updates when the training procedure is relaxed. As a remedy,
Wasserstein GANs~\citep{2017arXiv170107875A} (WGANs) reformulate the adversarial
setup in order to minimize the Wasserstein-1 distance $W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  )$
by replacing the adversarial classifier with a 1-Lipschitz adversarial critic
$d(\cdot; \bfphi) : \mathbb{R}^d \to \mathbb{R}$, such that
\begin{align}
& W( p_r(\mathbf{x} ), p(\mathbf{x} | \bftheta)  ) \nonumber \\
&= \sup_{\bfphi}  \mathbb{E}_{\tilde{\mathbf{x}}\sim p(\mathbf{x}|\bftheta)} [d(\tilde{\mathbf{x}}; \bfphi)] - \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x}; \bfphi)] \nonumber \\
&\triangleq \sup_{\bfphi} \mathcal{L}_W.
\end{align}
Under the WGAN-GP formulation of \citep{2017arXiv170400028G}
for stabilizing the optimization procedure,
training $d$ and $g$ results in alternating gradient updates on $\bfphi$ and $\bftheta$ in order to respectively minimize
\begin{align}
    {\cal L}_d =\,&  \mathcal{L}_W + \lambda \mathbb{E}_{\hat{\mathbf{x}} \sim p({\hat{\mathbf{x}})}} [(|| \nabla_{\hat{\mathbf{x}}} d({\hat{\mathbf{x}}};\bfphi) ||_2 - 1)^2] \\
    {\cal L}_g =\,& - \mathcal{L}_W
\end{align}
where ${\hat{\mathbf{x}}} := \epsilon \mathbf{x} +
(1-\epsilon)\tilde{\mathbf{x}}$, for $\epsilon \sim U[0,1]$, $\mathbf{x} \sim
p_r(\mathbf{x})$ and $\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)$.

@@

Variational optimization~\citep{2012arXiv1212.4507S,staines2013optimization} and evolution strategies~\citep{2011arXiv1106.4487W} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the observation that
\begin{equation}
    \min_{\bftheta} f(\bftheta) \leq \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] = U(\bfpsi),
\end{equation}
where $q(\bftheta|\bfpsi)$ is a proposal distribution with parameters $\bfpsi$ over input values $\bftheta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal distribution is flexible enough, the parameters $\bfpsi$
can be updated to place its mass arbitrarily tight around the optimum $\bftheta^* = \min_{\bftheta \in \Theta} f(\bftheta)$.

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound
$U(\bfpsi)$ is differentiable with respect to $\bfpsi$, and using the log-likelihood
trick its gradient can be rewritten as:
\begin{align}\label{eqn:approx-grad}
    \nabla_\bfpsi U(\bfpsi) &= \nabla_\bfpsi \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta)] \nonumber \\
    &= \nabla_\bfpsi \int f(\bftheta)  q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
   &= \int f(\bftheta) \nabla_\bfpsi q(\bftheta|\bfpsi)  d\bftheta \nonumber \\
   &= \int \left[ f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi) \right]  q(\bftheta|\bfpsi) d\bftheta \nonumber \\
    &= \mathbb{E}_{\bftheta \sim q(\bftheta|\bfpsi)} [f(\bftheta) \nabla_\bfpsi \log q(\bftheta|\bfpsi)]
\end{align}
Effectively, this means that provided that the score function $\nabla_\bfpsi \log
q(\bftheta|\bfpsi)$ of the proposal is known and that one can evaluate
$f(\mathbf{\bftheta})$ for any $\bftheta$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\bfpsi)$ with stochastic gradient descent (or a variant thereof, robust to noise
and parameter scaling).

@@

Algorithm~\ref{alg:avo} represents the simplest version of AVO; however, the
variance of the noisy estimator of the gradients may be too large to be useful in many problems.
We use Adam~\citep{2014arXiv1412.6980K}, but note the opportunity to use instead the Natural Evolution Strategy
algorithm~\citep{2011arXiv1106.4487W} or variance reduction techniques such as
control variates~\citep{2017arXiv171100123G} or finite differences~\citep{buesingstochastic}.

@@

The variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
the parametrized marginal distribution of the generated data
\begin{equation}\label{eq:marginal_likelihood}
\qxpsi = \int  p(\mathbf{x}|\bftheta) q(\bftheta|\bfpsi) d\bftheta.
\end{equation}
We can think of $q(\bfx|\bfpsi)$ as a \textit{variational program} as described
in \citep{2016arXiv161009033R}, though more complicated than a simple
reparametrization of normally distributed noise $\bfz$ through a differentiable
function. In our case, the variational program is a
marginalized, non-differentiable  simulator.  Its density is  intractable;
nevertheless, it can generate samples for $\bfx$ whose expectations are differentiable with
respect to $\bfpsi$.
Operationally, we sample from this marginal model via
\begin{equation}\label{eqn:p_psi}
\mathbf{x} \sim \qxpsi \triangleq \bftheta \sim q(\bftheta|\bfpsi), \bfz \sim p(\bfz|\bftheta), \mathbf{x} = g(\bfz; \bftheta).
\end{equation}

We can view the optimization of $q(\bfx | \bfpsi)$ with respect to $\bfpsi$ through the lens of empirical Bayes, where the data are used to optimize a prior within the family $q(\bftheta|\bfpsi)$. If $\rho$ is the KL distance, $\bfpsi^*$ would correspond to the maximum marginal likelihood estimator advocated by Rubin~\citep{rubin1984}.
When $\rho$ is the Wasserstein distance, $\bfpsi^*$ is referred to as the
minimum Wasserstein estimator (MWE).
When the model is well specified, the MWE
coincides with the true data-generating parameter; however, if the model is
misspecified, the MWE is typically different from the maximum likelihood
estimator (MLE).
Thus, if the simulator $p(\bfx |
\bftheta)$ is misspecified, $q(\bftheta | \bfpsi)$ will attempt to smear it so that the marginal
model $q(\bfx | \bfpsi)$ is closer to $p_r(\bfx)$. However, if the simulator is well specified,
then $q(\bftheta | \bfpsi)$ will concentrate its mass around the true data-generating parameter.

@@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{figures/poisson.pdf}
\caption{Discrete Poisson model with unknown mean.
 ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after training. For both $\gamma=0$ and $\gamma=3$, the distributions correctly concentrate their density around
            the true value $\log(\lambda^*)$. Entropic regularization ($\gamma=3$) results in a tighter density.
 ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parametrizations of the simulator closely reproduce the true distribution, with better results when enabling entropic regularization.
 ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
 }\label{fig:poisson}
\end{figure}

@@

The observed data is sampled from a Poisson with mean $\lambda^* = 7$.
Algorithm~\ref{alg:avo} is run for 3000 iterations with mini-batches of size $M=50$
and the following configuration. For the critic $d$, we use a 3-layer MLP with 10
hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
$n_\text{critic}=100$ iterations with a step size $\alpha=0.001$, with $\beta_1=0.5$ and $\beta_2=0.9$.
% Its inner first and second moment vectors are reset at
% each outer epoch in order to avoid building momentum in staled directions.
For
estimating $\lambda^*$, we parametrize $\bftheta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
$q(\bftheta|\bfpsi)$ initialized with a mean at $\log(5)$ and unit variance. At
each iteration, parameters $\bfpsi$ are updated by taking one Adam step and the same hyper-parameters as for the critic.
The gradient penalty coefficient is set to
$\lambda=0.001$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=3$.

The top left plot in Figure~\ref{fig:poisson} illustrates the resulting proposal
distributions $q(\bftheta|\bfpsi)$ after AVO.  For
both $\gamma=0$ and $\gamma=3$, the proposal distributions correctly concentrate
their density around the true parameter value $\log(\lambda^*) = 1.94$. Under
the effect of entropic regularization,
the proposal distribution for $\gamma=3$ concentrates its mass tightly,
yielding in this case precise inference.  The top right plot compares the
model distributions to the true distribution.  As theoretically expected from
adversarial training, we see that the resulting distributions align with
the true distribution, with in this case visually slightly better results for the penalized
model.  The bottom plot of Figure~\ref{fig:poisson} shows empirical estimates
of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=3$,
the curves fall towards $0$, which indicates that
$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)}
[d(\tilde{\mathbf{x}};\bfphi)] \approx \mathbb{E}_{\mathbf{x} \sim
p_r(\mathbf{x})} [d(\mathbf{x};\bfphi)]$ and that the critic cannot distinguish
between true and model data. This confirms that adversarial variational optimization
works despite the discreteness of the data and lack of access to the
density $p(\bfx | \bftheta)$ or its gradient.

@@

As a second illustrative example, we consider a generator producing
5-dimensional continuous data, as originally specified in Section 4.2 of
\citep{cranmer2015approximating}.
More specifically, we consider the following
generative process:
\begin{itemize}
\item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
$z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
$z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
$z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
$z_3 \sim {\text{Exponential}(\lambda=3)}$, and
$z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
\item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
semi-positive definite $5 \times 5$ matrix defining a fixed projection
of $\bfz$ into the observed space.
\end{itemize}
Again, AVO does not have access to the density or its gradient, only samples from the generative model.
We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored  (mean field) Gaussian
proposal distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component was
initialized with zero mean and unit variance.
Hyper-parameters are set to $M=50$, $n_\text{critic}=100$, $\lambda=0.01$, $\gamma=1$ and
Adam configured with a step size $\alpha=0.001$ with $\beta_1=0.5$ and $\beta_2=0.9$.
The architecture of the critic is the same as in the previous example.

@@

This work sits at the intersection of several lines of research related to
likelihood-free inference, approximate Bayesian computation (ABC),
implicit generative models, and variational inference.
Viewed from the literature around implicit generative models based on neural networks,
the proposed method can be considered as a direct adaptation of
generative adversarial networks~\citep{goodfellow2014generative,2017arXiv170104862A} to
non-differentiable simulators using variational optimization~\citep{2012arXiv1212.4507S}.
